\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-12-18 doc/publications/plos16/plos16.tex
added some figures
changes to plos16.tex

A tree based model is capable of taking this interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it it possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. {\sc CloudForest}~\cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest\ implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
of \cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.  Finally, given the role
different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
prediction accuracy of the model.

@@

\subsubsection{Synthetic data} 
\marginpar{check details of simulation. Piotr} Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.  The dichotomous response is generated from a Bernoulli trial with $p$ (probability of success)
calculated by applying an inverse logit transformation to the continuous linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 1.6, a_{100} = 1.8, a_{1000} = 2.2, a_{10000} = 2.0, a_{10000}=2.3$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.

@@

\subsubsection{simulation 1}

Piotr's simulation with $n=1000$ and $p= 10000$

\begin{figure}[tbhp] 
 \begin{adjustwidth}{-1.00in}{0in}
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
    \caption{OOB classification error} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \begin{subfigure}[b]{0.5\linewidth}
    \centering
    \includegraphics[totalheight=8cm]{./figs/rbo.png}
    \caption{The  rank-biased-overlap measure.} 
    \label{fig7:b} 
    \vspace{4ex}
  \end{subfigure} 
  \caption{{\bf Variable importance (Mean Decrease Gini) and OOB (out of bag) classification error estimate for random
      forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}  Each measure is shown
    as a function of \mtry and \ntree. {\bf a0} OOB (out of bag) classification error.  {\bf b)} The rank-biased-overlap \cite[]{Webber.et.al.2010}. }
  \label{figure:synth}
\end{adjustwidth}
\end{figure}

\FloatBarrier
\subsubsection{simulation 2}
Similar to Piotr's simulation with  except $y= w_1x_1 w_2x_2 w_3x_3 w_4x_4 w_5x_5$. $n=1000$ and $p= 10000$
The significant variables can no longer be recovered by single variable tests for correlation with $y$

@@

\begin{figure}[tbhp]
\includegraphics[totalheight=6cm]{/home/dun280/Dropbox/random_forests/exploration/figs/ranger.png}
\caption{ ranger}
\label{figure:ranger.png}
\end{figure}

>>>>

A tree based model is capable of taking this interaction of variables into account \cite{Wright.et.al.2016}. In addition, random forests are well
suited for processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have
the propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting.  Secondly, random
forests are also very easy to parallelize. As the forest is a sum of decision trees, it is possible to grow separate
trees on different processors and combine the trees. See the discussion of the properties of random forests in Section
\ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

In the first section we explore some of the properties of random forests as these will also be apparent in the
\cursedforest\ implementation. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability
of \cursedforest\ in respect to the dimensionality of the data, building a random forest model on whole-genome data from
the 1000 Genomes Project. We report the error in predicting ethnicity.  Finally, given the role
different parameter values can play in model construction, we explore the effect tuning these parameters can have on the
prediction accuracy of the model.

@@

\subsubsection{Synthetic data} 
Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
\begin{align}
y=& \prod_{i=1}^{i=5} {w_i x_i} + \epsilon \textrm{ for $x_i \neq 0$} 
\intertext{where}
\epsilon \sim & N(0, \sigma_\epsilon^2= 0.12)
\end{align}
The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
In this section we demonstrate two features of \cursedforest\, variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the method section.

@@

@@

\begin{figure}[tbhp]
\includegraphics[totalheight=6cm]{./figs/ranger.png}
\caption{ ranger}
\label{figure:ranger.png}
\end{figure}
